---
title: "Comparison_to_BB"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Part 1: Age

## Load the data
The age data will be compared first.

```{r loading data, warning= FALSE, message=FALSE}
library(dplyr)
library(qqman)
library(gplots)
library(pheatmap)
library(ggfortify)

res <- read.csv("res_f__chip_included_manifest.csv", header = T, row.names = 1)
bb <- read.csv("/mnt/data1/BrainData/Results/Age_Sex/PFC/LM_MLM_Age_Sex_CovariedChipNeuronalProportion_MLMethod.csv", header= T, row.names = 1)

epicManifest<-read.csv("/mnt/data1/IndiaWorkshop/epicManifest.csv", stringsAsFactors = F, header=T)
```

## Pull out the significant probes

``` {r sig probes}
res_age <- res[which(res$Age_P <= 0.05),]
bb_age <- bb[which(bb$Age.P <= 0.05),]

res_sex <- res[which(res$Sex_P <= 0.05),]
bb_sex <- bb[which(bb$Sex.P <= 0.05),]
```

## How many probes match

``` {r combining ages}
#find the probes in both dataset
bb_res_match <- rownames(res_age)[match(rownames(bb_age), rownames(res_age))]
bb_res_match <- as.data.frame(bb_res_match)
bb_res_match <-bb_res_match[which(bb_res_match$bb_res_match != "NA"),]
bb_res_match <- as.data.frame(bb_res_match)

#extrapolate the beta, sd and p values for these probes from both datasets
res_m <- res_age[which(bb_res_match$bb_res_match %in% rownames(res_age)),c(1,2,3,4,5,6)]
bb_m <- bb_age[which(bb_res_match$bb_res_match %in% rownames(bb_age)),c(1,2,3,5,6,7)]

bb_res_age <- merge(res_m, bb_m, by  = "row.names")
rownames(bb_res_age) <- bb_res_age[,1]
bb_res_age <- bb_res_age[,-1]

rm(bb_age,bb_m,bb_res_match, res, res_age, res_m)
res <- bb_res_age
```

## Add manifest data

```{r manifest}
epicManifest<-epicManifest[match(rownames(res), epicManifest$Name),]
res<-cbind(res, epicManifest[,c("CHR", "MAPINFO", "UCSC_RefGene_Name", "UCSC_RefGene_Group", "Relation_to_UCSC_CpG_Island", "Regulatory_Feature_Name", "Regulatory_Feature_Group", "DNase_Hypersensitivity_NAME", "OpenChromatin_NAME", "TFBS_Evidence_Count", "Methyl450_Loci", "SNP_ID", "SNP_DISTANCE", "SNP_MinorAlleleFrequency")])
```


## Compare the probes between UK and BB via
```{r load India data}
load("/mnt/data1/NIMHANS/Down_stream_analysis/NIMHAM_FPC_Normalised_snpsremoved.rdat") 
pheno_india <- pheno 
betas_india <- betas

rm(pheno, betas)

```

```{r load UK data}
load("/mnt/data1/BrainData/NormalisedData.Rdata")
pheno_uk <-  full.pheno
betas_uk <- betas.dasen

rownames(pheno_uk) <- pheno_uk$Sentrix_Full #set rownames as sentrix ids

#now we need to filter for frontal brain regions
pheno_uk <- filter(pheno_uk, pheno_uk$BrainRegion == "FC")
rownames(pheno_uk) <- pheno_uk$Sentrix_Full #set rownames as sentrix ids
betas_uk <- as.data.frame(betas_uk)
betas_uk <- select(betas_uk, one_of(rownames(pheno_uk)))
 
head(colnames(betas_uk) == pheno_uk$Sentrix_Full) #quickj check to see that the correct samples have been pulled from the uk data

rm(full.pheno, betas.dasen)

```


## Create metadata 
This metadata is to help us visualise the sample information when plotting the heatmap
```{r metadata}
cell_rows <- as.data.frame(pheno_india$Sex, rownames(pheno_india))
colnames(cell_rows)<- "Sex"
cells_rows2 <- as.data.frame(pheno_uk$Sex, rownames(pheno_uk))
colnames(cells_rows2)<- "Sex"
cell_row <- rbind(cell_rows, cells_rows2)
cell_age <- as.data.frame(pheno_india$Age, rownames(pheno_india))
colnames(cell_age)<- "Age"
cell_age$Age <- as.numeric(as.character(cell_age$Age))
cells_age2 <- as.data.frame(pheno_uk$Age, rownames(pheno_uk))
colnames(cells_age2)<- "Age"
cells_age2$Age <- as.numeric(as.character(cells_age2$Age))
cell_ages <- rbind(cell_age, cells_age2)
ind <- as.data.frame(rep('India',nrow(pheno_india)), rownames(pheno_india))
colnames(ind) <- 'Source'
eng <- as.data.frame(rep('UK',nrow(pheno_uk)), rownames(pheno_uk))
colnames(eng) <- 'Source'
loc <- rbind(ind,eng)

cells <- cbind(cell_row, cell_ages, loc)
```


```{r comparison of age}
#filter the probes that are significant in both datasets from raw matrices
betas_u <- betas_uk[which(rownames(betas_uk) %in% rownames(res)),]
betas_i <- betas_india[which(rownames(betas_india) %in% rownames(res)),]

#combine the two so that a single matrix with all samples is created
betas_ui <- merge(betas_u, betas_i, by = "row.names")
rownames(betas_ui) <- betas_ui$Row.names
betas_ui <- betas_ui[,-1]#remove rownames column
 
pheatmap(betas_ui, scale = 'row',
         annotation_col = cells,
         cluster_rows = F,
         show_rownames = F,
         show_colnames = F,
         main="Top significant age probes between UK and Indian samples (frontal)")

```


#Part 2: Sex

Since the data was already loaded frontal, we can pull out the sex information from the frontal beta values and UK population

```{r comparison of sex}
#take out the significant probes from beta matrix
bb_res_match <- rownames(res_sex)[match(rownames(bb_sex), rownames(res_sex))]
bb_res_match <- as.data.frame(bb_res_match)
bb_res_match <-bb_res_match[which(bb_res_match$bb_res_match != "NA"),]
bb_res_match <- as.data.frame(bb_res_match)

#extrapolate the beta, sd and p values for these probes from both datasets
res_m <- res_sex[which(bb_res_match$bb_res_match %in% rownames(res_sex)),c(4,5,6)]
bb_m <- bb_sex[which(bb_res_match$bb_res_match %in% rownames(bb_sex)),c(5,6,7)]

#merge the two datasets together and remove tables no longer required
bb_res_sex <- merge(res_m, bb_m, by  = "row.names")
rownames(bb_res_sex) <- bb_res_sex[,1]
bb_res_sex <- bb_res_sex[,-1]
rm(bb_sex,bb_m,bb_res_match, res, res_sex, res_m)
res2 <- bb_res_sex

#filter the probes that are significant in both datasets from raw matrices
betas_u2 <- betas_uk[which(rownames(betas_uk) %in% rownames(res2)),]
betas_i2 <- betas_india[which(rownames(betas_india) %in% rownames(res2)),]

#combine the two so that a single matrix with all samples is created
betas_ui2 <- merge(betas_u2, betas_i2, by = "row.names")
rownames(betas_ui2) <- betas_ui2$Row.names
betas_ui2 <- betas_ui2[,-1]#remove rownames column
 
identical(rownames(betas_ui2), rownames(cells))


pheatmap(betas_ui2, scale = 'row',
         annotation_col = cells,
         cluster_rows = F,
         show_rownames = F,
         show_colnames = F,
         main="Top significant sex probes between UK and Indian samples (frontal)")

```

#pca
To help us visualise how samples are related to each and whether there are any samples no grouped well.

```{r pca of age and sex}

betas_ui <- betas_ui[,order(colnames(betas_ui))]
cells <- cells[order(rownames(cells)),]
colnames(betas_ui) ==  rownames(cells)
beta_uit <- t(betas_ui)
pca_age <- prcomp(beta_uit, scale = T, center = T)

betas_ui2 <- betas_ui2[,order(colnames(betas_ui2))]
cells <- cells[order(rownames(cells)),]
colnames(betas_ui2) ==  rownames(cells)
beta_ui2t <- t(betas_ui2)
pca_sex <- prcomp(beta_ui2t, scale = T, center = T)


par(mfrow = c(2,2))
autoplot(pca_age, data = cells, colour = 'Source', 
         title= "Age pca for UK & India samples")
autoplot(pca_age, data = cells, colour = 'Age', 
         title= "Age pca for UK & India samples")
autoplot(pca_age, data = cells, colour = 'Sex', 
         title= "Age pca for UK & India samples")
screeplot(pca_age, npcs = 20, type =  "lines")




par(mfrow = c(2,2))
autoplot(pca_sex, data = cells, colour = 'Source', 
         title= "Sex pca for UK & India samples")
autoplot(pca_sex, data = cells, colour = 'Age', 
         title= "Sex pca for UK & India samples")
autoplot(pca_sex, data = cells, colour = 'Sex', 
         title= "Sex pca for UK & India samples")
screeplot(pca_sex, npcs = 20, type =  "lines")

```

#correlation of significant probe's methylation
To see that the significant probes have the same direction in India and UK datasets.
```{r pearsons}
library(ggpubr)

ggscatter(bb_res_age, x = 'Age_Beta', y = 'Age.Coeff',
          add = 'reg.line',
          add.params = list(color = 'blue', fill = 'lightgray'),
          conf.int = T) +
  stat_cor(method = 'pearson') +
  labs(title = "Correlation between UK and Indian significant sites on age (frontal)",
       x = 'India', y = 'UK')

ggscatter(bb_res_sex, x = 'Sex_Beta', y = 'Sex.Coeff',
          add = 'reg.line',
          add.params = list(color = 'blue', fill = 'lightgray'),
          conf.int = T) +
  stat_cor(method = 'pearson') +
  labs(title = "Correlation between UK and Indian significant sites on sex (frontal)",
       x = 'India', y = 'UK')

```


